---
title: "Links between openness to experience, imagery and creative problem solving"
author: "Jon May, Jessica Hart, Paul Sharpe and Andy Wills"
output:
  html_document:
    highlight: pygment
---

```{r setup, include=FALSE}
## DEVELOPERS: Uncomment one option, as appropriate
## Show only commands.
## knitr::opts_chunk$set(echo = TRUE, message = FALSE, results='hide', fig.keep = 'none', comment=NA)
## See also
## https://stackoverflow.com/questions/39709745/decreasing-the-line-thickness-and-corr-font-size-in-ggpairs-plot
## Show commands and output.
knitr::opts_chunk$set(echo = TRUE, comment=NA, cache = TRUE)
library(pander)
```

## Before you start

This is an advanced worksheet, which assumes you have completed the [Absolute Beginners' Guide to R](https://ajwills72.github.io/rminr/#beginners) course, the [Research Methods in Practice (Quantitative section)](https://benwhalley.github.io/rmip/overview-quantitative.html) course, and the [Intermediate Guide to R](https://ajwills72.github.io/rminr/#rmip) course. 

## Contents

- [Introduction](#intro)
- [Loading data](#load)
- [Correlation between openness and problems solved](#cor-o-problems)
- [Correlation matrix: other personality factors and problems solved](#cor-matrix)

<a name="intro"></a>

## Introduction

This worksheet describes a full analysis pipeline for an undergraduate student dissertation which explored relationships between personality, imagery and creative problem solving. Forty-eight students were tested to address three hypotheses. First, the researchers predicted that participants with more open personality types would be better at solving a selection of problems requiring creative solutions. Second, they predicated that participants with more vivid mental imagery would be better at solving the problems. Third, they predicted a relationship between divergent thinking and an ability to solve the problems.

Personality was measured using a simplified version of Costa and McCrae’s (1992) "big five" personality questionnaire. Mental imagery was measured using the PsiQ Plymouth Sensory Imagery Questionnaire (Andrade et al., 2013). Divergent thinking was measured using a 'flexible thinking task', which measured fluency, flexibility and originality. The problems requiring creative solutions were taken from May (1987).

<a name="load"></a>

## Loading data

Open the `rminr-data` project we used [previously](preproc.html#load).

Ensure you have the latest files by asking git to "`pull`" the repository. Select the `Git` tab, which is located in the row of tabs which includes the `Environment` tab. Click the `Pull` button with a downward pointing arrow. A window will open showing the files which have been pulled from the repository. Close the `Git pull` window. The `case-studies` folder should contain the folder `jon-may`.

Next, create a new, empty R script and save it in the `rminr-data` folder as `cs-jon-may.R`. Put all the commands from this worksheet into this file, and run them from there. Save your script regularly.

We start by loading the data.

```{r excel, message=FALSE}
rm(list = ls()) # clear the environment
library(tidyverse)
```

**Explanation of commands:**

We clear the workspace, and load `tidyverse` package, then read the four data files.

<a name="preproc"></a>

## Preprocessing

### Problems

Participants attempted to solve seven divergent thinking problems from a booklet. Each problem was coded as solved (=1) or not solved (=0) and the experiementers created a file with a column for the participant id (subj) and a column of 1s and 0s for each problem. 

```{r}
problems <- read_csv('case-studies/jon-may/problems.csv')
solved <- problems %>%
  rowwise() %>%
  mutate(problems = sum(c_across(bridges:dots))) %>%
  select(subj, problems)
```

**Explanation of commands:**
We read the problems CSV file into the dataframe `problems`. Then we use `rowwise` and `sum(c_across(...))` to add up the total problems solved for each participant, creating a new dataframe called `solved`, containing just the participant id `subj` and the total solved `problems`.

### OCEAN

The Big5 OCEAN questionnaire contained 50 items (ten for each factor), each answered 1 to 5, with some needing to be reverse scored. The response data was entered into a CSV file **big5** (the experimenters made one obvious data entry error). The items had been presented in an intermixed order, so to facilitate scoring, the experimenters also made a CSV file **oceankey** containing a scaleID for each column in **big5**, with the ten items for each scale being named with the initial letter of their scale and prefixed with **r** if they needed to be reversed (e.g., column `S36` was the ninth item of the Extraversion scale, and was reversed, so was named as ItemID `rE09`.

```{r}
ocean   <- read_csv('case-studies/jon-may/big5.csv')

oceankey<-read.csv("case-studies/jon-may/oceankey.csv")

scaleIDs <- oceankey$ScaleID                 # make a vector of the new variable names
names(scaleIDs) <- oceankey$ItemID           # name each new name with the old name so it can be looked-up

```
After reading in both of these files, a named vector `scaledIDs` is created. This allows column names to be 'looked-up', so e.g. `scaleIDs["S36"]` returns `rE09`, the scaleID for column S36.

```{r}
ocean.scales<-ocean %>%
   pivot_longer(S1:S50,names_to='Item', values_to='Value') %>%     # put the 50 items into a column
   mutate(ScaleID=scaleIDs[Item])%>%                               # lookup the new names from the vector
   select(-Item) %>%                                               # drop the columns we don't need
   mutate(Value=ifelse(Value>5,NA,ifelse(Value<1, NA, Value))) %>% # screen for absurd values
   mutate(Value=ifelse(grepl("r",ScaleID), 6-Value, Value)) %>%    # reverse score items containg 'r' ...
   mutate(ScaleID = sub("r", "", ScaleID)) %>%                     # ... and remove the 'r'
   pivot_wider(names_from="ScaleID", values_from="Value") %>%      # make the data wide again
   select(subj,sort(names(.))) %>%                                 # sort the columns by the new names
   rowwise() %>%                                                   # for each participant...
   mutate(openness=mean(c_across(o01:o10), na.rm = TRUE),          # ... mean of 10 items in each scale
          conscientiousness=mean(c_across(C01:C10), na.rm = TRUE),
          extraversion=mean(c_across(E01:E10), na.rm = TRUE),
          agreeableness=mean(c_across(A01:A10), na.rm = TRUE),
          neuroticism=mean(c_across(N01:N10), na.rm = TRUE)) %>%
   select(subj, openness, conscientiousness, extraversion, agreeableness, neuroticism)
```
**Explanation of commands**
This long chunk of tidy code renames all of the item, cleans up the data, reverse scores the items that need reversing, and computes means for each of the five factors.  

* First,  the `pivot_longer` command puts the values from all fifty columns into a single new column called `Value`, with the column names saved into `Item`. 

* The `mutate` command then creates a new column mapping the Item names into the more useful ScaleIDs (looking them up using the named vector), and  the Item column is deleted. 

* The three `mutate` commands first turn any value larger than the scale maximum to NA (this gets rid of the obvious data entry problem); then use `grepl("r", ScaleID)` to find any items whose names begin with "`r`" and subtracts them from 6 to reverse score them; and removes any "`r`" from the item names. 

* The `pivot_wider` command puts each participant's data into a single row again, now with the new r-less scaleID column names, and sorts them alphabetically, while keeping the `subj` column first. 

* Finally the `rowwise` and `mean(c_across(...))` commands create the scale means, and only these need to be retained in the final data frame called **ocean.scales**.

### Flexible thinking test

The ‘Flexible Thinking Test’ was a boolet with three subtests in it. Each subtest gave a score, and these were typed into a CSV file called ftt. They need to be added up

```{r}
ftt <- read_csv('case-studies/jon-may/ftt.csv') %>%
   rowwise() %>%
   mutate(ftt = sum(c_across(ftt1:ftt3))) %>%
   select(subj, ftt)
```
Instead of reading the file into a dataframe and then manipulating it, we here do it all in one tidy command to make a two column dataframe *ftt* with the participant ID and the FTT total.

### Vividness of Imagery
The PsiQ vividness of imagery scale has 35 items, five for each of seven modalities. For this analysis, we only want the first five items which measured vividness of visual imagery. These were answered on a scale running from 0 (not at all vivid) to 10 (as vivid as real life), and the score is the mean of all five items.

```{r}
psiq <- read_csv('case-studies/jon-may/psiq.csv') %>%
   rowwise() %>%
   mutate(psiq = mean(c_across(2:6))) %>%
   select(subj, psiq)
```
### Combining the dataframes
We now have four dataframes for problems solved, the Big 5, the FTT, and the PsiQ. We need to join these together using the participant ID `subj` as the key.

```{r}
data <- full_join(ocean.scales, solved, by = 'subj') %>%
   full_join(ftt, by = 'subj') %>%
   full_join(psiq, by = 'subj')
```
Using `full_join` makes sure that all of the data is retained, even if not all of the participants were in eall of the four data frames.

## Mean, Standard deviation, skewness, kurtosis for all independent variables (n=48)

## Histogram and curve showing distribution of problem scores (n=48)

https://stackoverflow.com/questions/6967664/ggplot2-histogram-with-normal-curve

```{r}
n_obs = sum(!is.na(data$problems))
bw = 1
mean <- mean(data$problems)
sd <- sd(data$problems)

data %>% ggplot(aes(problems))  + 
  geom_histogram(colour = "black", binwidth = bw) + 
  stat_function(fun = function(x) 
    dnorm(x, mean = mean, sd = sd) * bw * n_obs) +
  xlab('Problems solved') + ylab('Count (participants)')
```

## Scatterplot and best fit line, openness vs. problems solved

You learnt how to create a scatterplot and a best fit line in the [regression worksheet](https://benwhalley.github.io/rmip/regression.html).

<a name="cor-o-problems"></a>

## Correlation between openness and problems solved

The researchers predicted that participants with more open personality types would be better at solving the problems. In other words, we should expect a positive correlation between openness and problem solving. We test a directional hypothesis with a one-tailed correlation. See the [More on relationships, part 2.](#corr-extended.html) worksheet for more details.

```{r}
cor_o_problems <- cor.test(data$problems, data$openness, alternative='greater')
cor_o_problems

# note that this is a 2-tailed test
library(BayesFactor)
cor_o_problems_bf <- correlationBF(data$problems, data$openness)
```

**Explanation of commands:**

We use a Pearson correlation to look at the relationship between openness and the number of problems solved. We specify `alternative='greater'` to indicate that we are predicting a positive correlation (one-tailed test). We also calculate a Bayes factor to test the evidence for the correlation.

**Explanation of output:**

Contrary to the first hypothesis, there is no evidence for a positive correlation  between openness and creative problem solving  (_r_ = `r round(cor_o_problems$statistic,2)`, one-tailed, _BF_ = `r round(BayesFactor::extractBF(cor_o_problems_bf)$bf,2)`).

<a name="cor-matrix"></a>

## Correlation matrix: other personality factors and problems solved

Correlation matrices are covered in the [Better tables](better-tables.html#cor-matrix) worksheet.

```{r cor-matrix, R.options = list(width = 10000)}
library(apaTables)
apa.cor.table(data %>% select(conscientiousness:problems), filename='table1.doc', table.number = 1)
```

**Explanation of output:**

There were no correlations between the other four personality factors and problem solving. Note that these are 2-tailed correlations, as it's not possible to specify 1-tailed tests with `apa.cor.table()`.


## Correlation between problems solved and flexible thinking

TODO: one-tailed

## Correlation between problems solved and vividness

TODO: one-tailed

## Pairs plot

Pairs plots are are covered in the [Better graphs](better-graphs.html#pairs) worksheet.

```{r pairs, class.source = 'numberLines lineAnchors', message=FALSE}
library(GGally)
source('themeapa.R')
data %>%
  select(psiq, openness, problems, ftt) %>%
  ggpairs(lower=list(continuous='smooth')) +
  theme_APA
```

## References

Andrade, J., May, J., Deeprose, C., Baugh, S.-J., & Ganis, G. (2014). [Assessing vividness of mental imagery: The Plymouth Sensory Imagery Questionnaire](https://onlinelibrary.wiley.com/doi/pdf/10.1111/bjop.12050?casa_token=-IMvMthFsKQAAAAA:_b6s7nwZ8JS-5fWM4Xxgey5sBBPpbIRPUupgZ2XFQ-jNYGNOK0ARomtRNgCIWUfZqUD7w9PBoLbeng) British Journal of Psychology, 105(4), 547–563. 

Costa, P. T., & McCrae, R. R. (1992). Neo personality inventory-revised (NEO PI-R). 
Psychological Assessment Resources Odessa, FL.

May, J. (1987). The cognitive analysis of flexible thinking. Unpublished PhD
thesis, University of Exeter.

___

This material is distributed under a [Creative Commons](https://creativecommons.org/) licence. CC-BY-SA 4.0. 

